From Reversible Programs to \\ Univalent Universes and Back

### 2017-07-02 upi.tex
fixed a few typos etc

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the question of when two programs are ``equivalent.'' Consider
the follwing six programs of type~$\bt \leftrightarrow \bt$

@@

\noindent We neither expect, nor desire, that all these programs of
the same type to be be equivalent. In the above, each group of
three does represent ``the same'' program but the two groups
represent distinct programs. Thinking extensionally, i.e., by looking
at all possible input-output pairs, it is indeed easy to verify this
fact, but we can do better. We can provide \emph{evidence} (i.e., a
program that manipulates programs) that can constructively transform
every program to an equivalent one. We show such a level-2 program
proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for
$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
the reader to map the steps below to manipulations on the diagram that
would incrementally simplify it:

@@

The second notion we need
is that of equivalence. We say that two types or spaces $A$ and $B$
are equivalent if there exist functions $f : A \rightarrow B$ and
$g : B \rightarrow A$ that compose to the identity function%
\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an
equivalence also requires a certain coherence between the compositions.}.
A notable point is that equivalences do not come with an induction principle.
We say a universe $\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.

@@

While univalent universes came first, it was later realised that
the more general univalent type families are more convenient to
work with, and allow the constructions of custom universes. 
We thus need to build up some definitions for what this means.
Note that below we will use ``point'' and ``member of a type'',
``space'' and ``type'', ``path'' and ``equivalence'' as
synonyms.  Furthermore our setting is an intuinistic type
theory (with no classical axioms such a choice or excluded
middle added); this implies that all mappings are
\emph{continuous}.

@@

As is often the case when confronted with new definitions, it is
instructive to disect \emph{non-examples}, as a means to understand
what ``goes wrong''.  As a first example of a non-univalent fibration,
consider $P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P =
\lambda\_.\mathbb{2}$.

>>>>

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the question of when two programs are ``equivalent.'' Consider
the following six programs of type~$\bt \leftrightarrow \bt$

@@

\noindent We neither expect, nor desire, that all these programs of
the same type to be equivalent. In the above, each group of
three does represent ``the same'' program but the two groups
represent distinct programs. Thinking extensionally, i.e., by looking
at all possible input-output pairs, it is indeed easy to verify this
fact, but we can do better. We can provide \emph{evidence} (i.e., a
program that manipulates programs) that can constructively transform
every program to an equivalent one. We show such a level-2 program
proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for
$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
the reader to map the steps below to manipulations on the diagram that
would incrementally simplify it:

@@

The second notion we need is that of equivalence. We say that two types or
spaces $A$ and $B$ are equivalent if there exist functions $f : A \rightarrow B$
and $g : B \rightarrow A$ that compose to the identity function.
\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an equivalence
  also requires a certain coherence between the compositions.}  A notable point
is that equivalences do not come with an induction principle.  We say a universe
$\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of equivalences
$\AgdaFunction{A} ≃ \AgdaFunction{B}$.

@@

While univalent universes came first, it was later realized that
the more general univalent type families are more convenient to
work with, and allow the constructions of custom universes.
We thus need to build up some definitions for what this means.
Note that below we will use ``point'' and ``member of a type'',
``space'' and ``type'', ``path'' and ``equivalence'' as
synonyms.  Furthermore our setting is an Intuitionistic type
theory (with no classical axioms such a choice or excluded
middle added); this implies that all mappings are
\emph{continuous}.

@@

As is often the case when confronted with new definitions, it is
instructive to dissect \emph{non-examples}, as a means to understand
what ``goes wrong''.  As a first example of a non-univalent fibration,
consider $P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P =
\lambda\_.\mathbb{2}$.
